# Layers: STACK

Total points: 46

## Introduction

You've already learned the basics of ggplot2: how to create scatterplots and boxplots, how to map aesthetics like color and shape, and how to customize your plots with themes and labels. 

In this chapter, we'll dive deeper into the **grammar of graphics** - the idea that you can build every graph from the same components: a dataset, a coordinate system, and geoms (geometric objects that represent data points). We'll explore:

- A deeper understanding of **aesthetic mappings** (inside vs outside aes)

- More **geometric objects** (geoms) for different plot types

- **Faceting** - creating multiple plots split by categories

- **Position adjustments** - how geoms are positioned relative to each other

We won't cover every single function and option, but we'll walk you through the most important and commonly used functionality in ggplot2.


### Prerequisites

Load the necessary packages into your session:

```{r}
library(tidyverse)                # for ggplot2 and mpg
library(patchwork)                # for combining plots
library(scales)                   # for axis formatting
```

## Aesthetic mappings

The `mpg` data frame is bundled with the ggplot2 package and contains
`r nrow(mpg)` observations on `r mpg |> distinct(model) |> nrow()` car
models from [fueleconomy.gov](https://fueleconomy.gov/). 
```{r include = F}
?mpg
names(mpg)
```

Let's start by visualizing the relationship between `displ` and `hwy`
for various `classes` of cars. We can do this with a scatterplot where
the numerical variables are mapped to the `x` and `y` aesthetics and the
categorical variable is mapped to an aesthetic like `color` or `shape`.
```{r}
# Mapping class to color works well
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

Notice how each class gets a different color automatically. Now let's try using shape instead:
```{r}
# Mapping class to shape produces warnings
ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```

When `class` is mapped to `shape`, we get two warnings:

> 1: The shape palette can deal with a maximum of 6 discrete values because more than 6 becomes difficult to discriminate; we have 7.
> Consider specifying shapes manually if you must have them.
>
> 2: Removed 62 rows containing missing values (`geom_point()`).

Since `ggplot2` will only use six shapes at a time, by default, additional groups will go unplotted when you use the shape aesthetic.
The second warning is related -- there are 62 SUVs in the dataset and they're not plotted.

We have to remember that ggplot2 does not replace, but layers. For example:
```{r}
# Create a saved plot
p1 <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) + 
  geom_point(size = 2)

p1  # p1
```

### Step1: make the points bigger
What if I wanted to make all the points bigger? If I add another geom_point(size = 6), it looks like it worked!
```{r}

p2 = p1 + geom_point(size = 6)

p2
```

### Step 2: Now try to make them smaller  

What if I wanted to make all the points smaller?
```{r}
p3 = p2+ geom_point(size = 2, color = "black") # "Wait... now I see BOTH layers! Small points on top of the original size!"

p3
```

**The grammar of graphics works like painting; each `+` adds a new layer of paint on top.** You can paint over something, but the old layer is still there underneath. When the new layer is bigger, it hides the original.

**Layers stack, they don't replace!**

To actually change `p1`, you need to recreate it from scratch:
```{r}
p1 <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 6)

p1
```

However, you CAN modify **non-data elements** - these are like changing the frame around your painting, not the painting itself:
```{r}
# Change the labels
p1 + labs(title = "Fuel Efficiency by Car Class")

# Change the theme  
p1 + theme_classic()
```

These don't add new data layers - they modify how the existing plot is displayed.

**The rule:**

- **Data layers** (geom_point, geom_line, etc.) → Stack like layers of paint

- **Display elements** (labs, themes) → Modify the frame and presentation

Sometimes it's helpful to compare plots side-by-side. The `patchwork` package makes this easy:
```{r}
# Save plots to variables
p1 <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()

p2 <- ggplot(mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()

# Combine side-by-side with +
p1 + p2
```

You can also stack plots vertically using `/`:
```{r}
# Stack plots vertically
p1 / p2
```

This is particularly useful when comparing different visualizations of the same data, as you'll often need to do in your research when comparing treatment groups or experimental conditions.

**What about other aesthetics?** Similarly, we can map `class` to `size` or `alpha` aesthetics as well, which control the size and the transparency of the points, respectively.
```{r}
# class mapped to size
ggplot(mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point() 
```
```{r}
# class mapped to alpha
ggplot(mpg, aes(x = displ, y = hwy, alpha = class)) +
  geom_point() 
```

Both of these produce warnings as well:

> Using size for a discrete variable is not advised.
> Using alpha for a discrete variable is not advised.

Mapping an unordered discrete (categorical) variable (`class`) to an ordered aesthetic (`size` or `alpha`) is generally not a good idea because it implies a ranking that does not in fact exist.

### Setting Visual Properties

Once you map an aesthetic, `ggplot2` automatically:

- Chooses appropriate values (colors, shapes, sizes)

- Creates a legend showing what each value means

- For x and y aesthetics, creates axis labels instead of legends

You can also set the visual properties of your `geom` manually as an argument of your `geom` function (*outside* of `aes()`) instead of relying on a variable mapping to determine the appearance. For example, we can make all of the points in our plot blue:
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = "blue") 
```

::: {.infobox .tip}
### Where do aesthetics go in ggplot2?

**The key question:** Is this value from your data, or is it a constant?

**From your data** (a column name) → INSIDE `aes()`
```{r}
# color varies by class (class is a column in mpg)
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

**A constant** (same for all points) → OUTSIDE `aes()`, in the geom layer
```{r}
# all points are pink
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "pink", size = 3)
```

**Memory trick:** Can you do `mpg$class` in R? YES → inside `aes()`  
Can you do `mpg$pink`? NO → outside `aes()`

**Important:** Constants must go in `geom_*()`, NOT in `ggplot()`:
```{r, eval=FALSE}
# ❌ This fails!
ggplot(mpg, aes(x = displ, y = hwy), color = "blue") +
  geom_point()

# ✅ This works!
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```
:::

### The Universal Rule for ALL Aesthetics

**Where can variables and constants go?**

| Location | Variable (from data)<br/>Example: `color = class` | Constant (fixed value)<br/>Example: `color = "blue"` |
|----------|---------------------|------------------------|
| `ggplot(aes(...))` | ✅ YES<br/>`aes(color = class)` | ❌ NO |
| `geom_*(aes(...))` | ✅ YES<br/>`aes(color = class)` | ❌ NO |
| `geom_*(...)` outside aes | ❌ NO | ✅ YES<br/>`color = "blue"` |


So far we have discussed aesthetics that we can map or set in a scatterplot, when using a point `geom`. You can learn more about all possible aesthetic mappings in the aesthetic specifications vignette at <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html>.

The specific aesthetics you can use for a plot depend on the `geom` you use to represent the data.

In the next section we dive deeper into `geoms`.

#### 1.2.1 Exercises {.unnumbered}

1. Create a scatterplot of `hwy` vs. `displ` where the points are pink filled in triangles. +2pts

::: {.fold .s .o}

```{r Ans-1.2.1.1}
ggplot(mpg, aes(x = hwy, y = displ)) +
  geom_point(color = "pink", shape = 24, size = 3, fill = "pink")
```
:::

2. What does the `stroke` aesthetic do? What shapes does it work with? (Hint: use `?geom_point`). Add stroke to the above plot. +2pts

::: {.fold .s .o}

**Ans-1.2.1.2:**
```{r Ans-1.2.1.2}
# Stroke controls the width of the edge/border for the shapes (21-25)
# that have a fill aesthetic.
ggplot(mpg, aes(x = hwy, y = displ)) +
  geom_point(color = "black", shape = 24, size = 3, fill = "pink", stroke = 5)
```
:::

3. Why did the following code not result in a plot with blue points? +2pts
```{r, eval = FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, color = "blue"))
```

::: {.fold .s .o}

**Ans-1.2.1.3:**

The points are NOT blue because `color = "blue"` is inside `aes()`.

When you put a value inside `aes()`, ggplot2 thinks it's a column name from your data. Since there's no column called "blue", ggplot2 creates a fake categorical variable with one value (the string "blue") and assigns it the first default color (salmon/coral).

**The fix - move color OUTSIDE aes():**
```{r}
# Correct: color is a constant, so it goes outside aes()
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy), color = "blue")
```
:::

4. What's wrong with the following code? How would you fix it? +2pts
```{r, eval = FALSE}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = class, size = 3)
```



::: {.fold .s .o}

**Ans-1.2.1.4:**

This code will give an error: `Error: object 'class' not found`

`class` is a column name from the data, so it must go **inside** `aes()`. When you put it outside, R looks for an object called `class` in your environment (not in the data) and doesn't find it.

**The fix:**
```{r}
# Correct: class is from data, so it goes inside aes()
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class), size = 3)
```

Note that `size = 3` stays outside `aes()` because it's a constant (all points are size 3).

**The key difference:**

- **Inside `aes()`**: R looks for columns in your data

- **Outside `aes()`**: R looks for objects in your environment
:::

5. What happens if you map an aesthetic to something other than a variable name, like `aes(color = displ < 5)`? Try it by adding color to the following plot: +2pts
```{r, eval = FALSE}
ggplot(mpg) + geom_point(aes(x = displ, y = hwy))
```

::: {.fold .s .o}

**Ans-1.2.1.5:**
```{r}
# It creates a logical variable with values `TRUE` and `FALSE` 
# for cars with displacement values below and above 5. 

# Notice color is inside aes() because we're creating a variable
# from the data (the result of displ < 5).

ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, color = displ < 5)) 
```
:::

6. What's wrong with the following code? Why doesn't it make blue points? How would you fix it? +2pts
```{r, eval = FALSE}
ggplot(mpg, aes(x = displ, y = hwy), color = "blue") + 
  geom_point()
```

::: {.fold .s .o}

**Ans-1.2.1.6:**

This code produces a warning: 
```
Warning message:
In fortify(data, ...) : Arguments in `...` must be used.
x Problematic argument:
* color = "blue"
```

The problem is that `color = "blue"` is in the **wrong place**. The `ggplot()` function only accepts `data` and `mapping = aes()` arguments. You cannot put styling constants (like `color`, `size`, `alpha`) directly in `ggplot()`.

Constants must go in the `geom_*()` layer, outside `aes()`:
```{r}
# Correct: constant in geom layer
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = "blue")
```

**Where things can go:**

- **In `ggplot()`**: Only `data` and `aes()` mappings

- **In `geom_*()` outside `aes()`**: Constants like `color = "blue"`, `size = 3`

- **In `aes()` (anywhere)**: Variables from your data like `color = class`

:::

## Geometric objects

You've already learned about `geom_point()` and `geom_boxplot()` in your previous lessons. A `geom` is the geometrical object that a plot uses to represent data. People often describe plots by the type of geom that the plot uses. For example, 

- bar charts use bar geoms (`geom_bar()`)

- line charts use line geoms (`geom_line()`)

- histograms use histograms geoms (`geom_histogram()`)

- density plots use density geoms (`geom_density()`). 

How are these two plots similar?

```{r}
# Scatterplot
ggplot(mpg, aes(x = displ, y = hwy)) +  
  geom_point() 
```

```{r}
# Smooth line
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_smooth()
```

Both plots contain the same x variable, the same y variable, and both describe the same data. But the plots are not identical. Each plot uses a different geometric object, `geom`, to represent the data. The first plot uses the point `geom`, and the second uses the smooth `geom`, a smooth line fitted to the data.

To change the `geom` in your plot, change the `geom` function that you add to `ggplot()`.

Every `geom` function in ggplot2 takes a `mapping` argument, either defined locally in the `geom` layer or globally in the `ggplot()` layer.

However, not every aesthetic works with every `geom`. You could set the shape of a point, but you couldn't set the "shape" of a line. If you try, ggplot2 will silently ignore that aesthetic mapping. On the other hand, you *could* set the `linetype` of a line.

`geom_smooth()` will draw a different line, with a different `linetype`, for each unique value of the variable that you map to `linetype`.

```{r}
# Shape doesn't work with geom_smooth - no legend appears
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_smooth()
```

```{r}
# Linetype DOES work with geom_smooth
ggplot(mpg, aes(x = displ, y = hwy, linetype = drv)) + 
  geom_smooth()
```

Here, `geom_smooth()` separates the cars into three lines based on their `drv` value, which describes a car's drive train. Here, `4` stands for four-wheel drive, `f` for front-wheel drive, and `r` for rear-wheel drive.

**As long as the aesthetic in the `ggplot` argument is compatible with the `geoms` used you are ok.**

**Notice there is no legend for the first plot because geom_smooth does not have a shape aesthetic.**

We can make this clearer by overlaying the lines on top of the raw data and then coloring everything according to `drv`.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() + 
  geom_smooth(aes(linetype = drv))
```

Notice that this plot contains **two geoms in the same graph**. This is a powerful feature of ggplot2 - you can layer multiple geoms to show different aspects of your data.

You can use the same idea to specify different `data` for each layer.
Here, we use red points to highlight two-seater cars.
The local data argument in `geom_point()` overrides the global data argument in `ggplot()` for that layer only.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_point(
   data  = mpg |> filter(class == '2seater'),
    color = "red",
    size = 3
  )
```

`geoms` are the fundamental building blocks of `ggplot2`. You can completely transform the look of your plot by changing its geom, and different `geoms` can reveal different features of your data. 

For example, these three plots show the same data (highway mileage) but reveal different patterns:

```{r}
# Histogram shows distribution
ggplot(mpg, aes(x = hwy)) + 
  geom_histogram(binwidth = 2)
```

```{r}
# Density plot shows smooth distribution
ggplot(mpg, aes(x = hwy)) + 
  geom_density()
```

```{r}
# Boxplot shows summary statistics and outliers
ggplot(mpg, aes(x = hwy)) + 
  geom_boxplot()
```

The histogram and density plot reveal that the distribution of highway mileage is bimodal and right skewed, while the boxplot reveals two potential outliers.

`ggplot2` provides more than 40 `geoms` but these don't cover all possible plots one could make. The best place to get a comprehensive overview of all of the `geoms` `ggplot2` offers is the [reference page:](https://ggplot2.tidyverse.org/reference).
To learn more about any single `geom`, use the help (e.g., `?geom_smooth`).

### Dealing with Overplotting

When you have many data points, they can overlap and hide each other. `geom_jitter()` adds a small amount of random noise to each point to spread them out:
```{r}
# Compare these two plots
# Points overlap - hard to see how many there are
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_point()

# Jittered points - easier to see individual observations
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_jitter(width = 0.2)
```

The `width` argument controls how much horizontal jitter to add. Try different values to see what works best!

`geom_jitter` is most often used with `geom_boxplot` which allows you to see all the points, not just the outliers.


```{r}
 # Jittered points - easier to see individual observations
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot(outlier.shape = NA) +  # Don't show outliers (we have jittered points)
  geom_jitter(aes(color = class), width = 0.2)


# Notice if you change the order of `geom_point` with `geom_jitter` the boxplot argument dominates (layers STACK)
ggplot(mpg, aes(x = class, y = hwy,color=class)) +
  geom_jitter(width = 0.2)+geom_boxplot(aes(color=class))


```


#### 1.3.1 Exercises {.unnumbered}

1. What `geom` would you use to draw a line chart? A boxplot? A histogram? A scatterplot with points spread out to avoid overlap? ('jitter plot') +4pts or +6pts (bonus point for line charts using TWO `geoms`, how do they differ?)

::: {.fold .s .o}

**Ans-1.3.1.1:**

- For a line chart you can use `geom_path()` or `geom_line()`. 
                    
- For a boxplot you can use `geom_boxplot()`.

- For a histogram, `geom_histogram()`.

- For a jitter plot, `geom_jitter()` 

:::

2. Create a plot showing highway fuel economy (`hwy`) by car class (`class`) that includes both a boxplot and jittered individual points. Make the points semi-transparent. +3pts

::: {.fold .s .o}

**Ans-1.3.1.2:**
```{r}
ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot(outlier.shape = NA) +  
  # Don't show outliers twice
  geom_jitter(width = 0.2, alpha = 0.5)
```

Note: `outlier.shape = NA` removes the outlier points from the boxplot so they only appear once (as jittered points) rather than twice.
:::

3. What does `show.legend = FALSE` do in the code below? What happens if you remove it? +2pts
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv), show.legend = FALSE)
```

::: {.fold .s .o}

**Ans-1.3.1.3:**

It removes the legend for the `geom` it's specified in. In this case it removes the legend for the smooth lines that are colored based on `drv`.

If you remove `show.legend = FALSE`, the legend will appear showing the three drive types (4, f, r) with their corresponding colors.
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv))
```

:::

4. What does the `se` argument to `geom_smooth()` do? +2pts

::: {.fold .s .o}

**Ans-1.3.1.4:**

It displays the confidence interval (standard error) around the smooth line.
You can remove this with `se = FALSE`.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(se = FALSE)
```
:::

5. Will these two graphs look different? Why/why not? +2pts

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

::: {.fold .s .o}

**Ans-1.3.1.5:**

No, they will look the same. In the first plot, the data and aesthetic mappings are defined globally in `ggplot()` and inherited by both `geoms`. In the second plot, each `geom` has its own local data and mappings that are identical. The result is the same plot.
:::

## Facets

Facets are a powerful way to split your plot into multiple subplots based on the values of one or more categorical variables. This is especially useful when you have **too many categories to distinguish by color alone**, or when you want to compare patterns across groups side-by-side.

### Faceting by One Variable

To facet your plot by a single variable, use `facet_wrap()`. The first argument of `facet_wrap()` is a formula (created with `~` followed by a variable name). The variable passed to `facet_wrap()` should be categorical.
```{r}
# Instead of trying to use 7 different colors for class...
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, nrow = 2)
```

This makes it much easier to see the relationship between engine size and fuel efficiency within each vehicle class. Notice how 2seaters behave differently from SUVs!

### Faceting by Two Variables

To facet your plot with the combination of two variables, use `facet_grid()`. The first argument of `facet_grid()` is also a formula, but now it's a double sided formula: `rows ~ cols`.
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
geom_point() +
facet_grid(drv ~ cyl)

```

The first variable (`drv`) determines the faceting by row, the second (`cyl`) by column. This lets you see how both drive type and number of cylinders affect the relationship between engine size and highway mileage.

### Adjusting Scales

By default each of the facets share the same scale and range for x and y axes. This is useful when you want to compare data across facets, but it can be limiting when you want to visualize the relationship within each facet better.

Setting the `scales` argument in a faceting function to `"free"` will allow for different axis scales across both rows and columns, `"free_x"` will allow for different scales across columns, and `"free_y"` will allow for different scales across rows.
```{r}

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
geom_point() +
facet_grid(drv ~ cyl,scales="free_y")


```

Notice how the y-axis now adjusts for each row, making it easier to see the variation in highway mileage within each drive type.

#### 1.4.1 Exercises {.unnumbered}

1. What happens if you facet on a continuous variable? Try faceting the plot below on `hwy`. +2pts

```{r}
ggplot(mpg, aes(x = drv, y = cyl)) + 
  geom_point()
```

::: {.fold .s .o}

**Ans-1.4.1.1:**

Faceting by a continuous variable results in one facet for each unique value of the continuous variable. We can see this in the scatterplot below of `cyl` vs. `drv`, faceted by `hwy`.

```{r}
ggplot(mpg, aes(x = drv, y = cyl)) + 
  geom_point() +
  facet_wrap(~ hwy)
```

This creates many facets (one for each unique highway mpg value), which usually isn't useful!
:::

2. Run the following code and facet on `drv` vs `cyl` using `facet_grid()`. What do the empty cells mean? +2pts

```{r}
ggplot(mpg, aes(x = drv, y = cyl)) + 
  geom_point()
```

::: {.fold .s .o}

**Ans-1.4.1.2:**

There are no cars with certain combinations of drive train and cylinders. For example, there are no cars with rear-wheel drive and 4 or 5 cylinders. Therefore the facet corresponding to that combination is empty. In general, empty facets mean no observations fall in that category.

```{r}
ggplot(mpg, aes(x = drv, y = cyl)) + 
  geom_point() +
  facet_grid(drv ~ cyl)
```
:::

3. What plots does the following code make? What does `.` do? +2pts

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ .)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ cyl)
```

::: {.fold .s .o}

**Ans-1.4.1.3:**

In the first plot, with `facet_grid(drv ~ .)`, the period means "don't facet across columns" - so you get facets in rows only.

In the second plot, with `facet_grid(. ~ cyl)`, the period means "don't facet across rows" - so you get facets in columns only.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ .)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(. ~ cyl)
```
:::

4. What's redundant about this plot? Why does `cyl` produce a gradient color scale instead of distinct colors? How could you improve the plot? +3pts
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = cyl)) +
  geom_point() +
  facet_grid(. ~ cyl)
```

::: {.fold .s .o}

**Ans-1.4.1.4:**

**Two issues here:**

**1. Redundancy:** Coloring by `cyl` when you're already faceting by `cyl` is redundant - the facet labels already tell you which cylinder count each panel represents. It's like putting up a sign that says "CATS" and then coloring all the cats orange to show they're cats. The color should show us something *different*!

**2. Gradient colors:** `cyl` produces a gradient because it's stored as a numeric/integer variable. When ggplot2 sees a numeric variable mapped to color, it creates a continuous color scale. But cylinder counts are really **categories** (4-cyl, 5-cyl, 6-cyl, 8-cyl), not a continuous measurement.

**Better approach - fix both issues:**
```{r}
# Use color for additional information (drive type)
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  facet_grid(. ~ cyl)
```

Or if you really want to color by cylinder count in a different context, convert to factor:
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = factor(cyl))) +
  geom_point()
```

**The principle:** Each visual element should show DIFFERENT aspects of your data. Don't waste an aesthetic repeating information!

:::

5. Take the first faceted plot in this section:

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class)
```

What are the advantages to using faceting instead of the color aesthetic? What are the disadvantages? +4pts

::: {.fold .s .o}

**Ans-1.4.1.4:**

**Advantages of faceting:**

- Each class is shown separately, with no overplotting

- Patterns within each class are easier to see

- Works well with many categories (color gets hard to distinguish with 7+ categories)

**Disadvantages of faceting:**

- Harder to compare classes directly to each other when they're in separate plots

- Takes up more space on the page

- Can be harder to see the overall pattern across all data

**Comparison:**
```{r}
# Faceting - easier to see within-class patterns
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ class, nrow = 2)

# Color - easier to compare across classes
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

You can also combine both approaches!
:::

6. Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? Why doesn't `facet_grid()` have `nrow` and `ncol` arguments? +2pts

::: {.fold .s .o}

**Ans-1.4.1.5:**

`nrow` sets the number of rows in the facet layout.
`ncol` sets the number of columns in the facet layout.

`facet_grid()` doesn't have `nrow` and `ncol` arguments because the number of rows and columns is determined by the number of unique values of the variables you're faceting by. For example, if you facet by `drv ~ cyl`, the number of rows is determined by the number of unique `drv` values (3), and the number of columns by the number of unique `cyl` values (4).
:::

## Position adjustments

Position adjustments control how geoms are arranged when they might overlap. This is particularly important for bar charts.

You can color bar charts using either the `color` aesthetic (outline) or the `fill` aesthetic (inside):
```{r}
# Fill colors the inside of the bars
ggplot(mpg, aes(x = drv, fill = drv)) + 
  geom_bar()
```

When you map `fill` to another variable, the bars are automatically **stacked**. Each colored rectangle represents a combination of the two variables:
```{r}
# Stacked bars (default)
ggplot(mpg, aes(x = drv, fill = class)) + 
  geom_bar()
```

The stacking is controlled by the `position` argument. You can use different positions to change how the bars are displayed:

#### `position = "fill"` - Show proportions

Makes each set of stacked bars the same height, making it easier to compare proportions across groups:
```{r}
# All bars go to 100%
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "fill")
```

#### `position = "dodge"` - Side-by-side comparison

Places bars directly beside one another, making it easier to compare individual counts:
```{r}
# Side-by-side bars
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "dodge")
```

**Position adjustments summary:**
- `position = "stack"` (default for bars): Shows total counts with composition breakdown
- `position = "fill"`: Shows proportions (all bars reach 100%)
- `position = "dodge"`: Shows side-by-side for easy comparison
- `position = "jitter"`: Adds random noise to points to avoid overplotting (you learned this with `geom_jitter()` earlier!)

To learn more about position adjustments, see `?position_dodge`, `?position_fill`, `?position_jitter`, and `?position_stack`.

#### 1.6.1 Exercises {.unnumbered}

1. What's the difference between `position = "stack"` and `position = "fill"` for bar charts? When would you use each? +2pts

::: {.fold .s .o}

**Ans-1.6.1.1:**

`position = "stack"` shows actual counts - bars have different heights based on total observations in each category. It is the default.

`position = "fill"` shows proportions - all bars are normalized to the same height (100%), making it easy to compare percentages across groups.

**Use stack when:** You want to see both totals and composition  
**Use fill when:** You want to compare proportions across groups
```{r}
# Stack: different heights (shows totals)
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "stack")

# Fill: same heights (shows proportions)
ggplot(mpg, aes(x = drv, fill = class)) +
  geom_bar(position = "fill")
```

:::

2. Create a dodged bar chart showing the count of cars by `class` and `drv`. +2pts

::: {.fold .s .o}

**Ans-1.6.1.2:**
```{r}
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "dodge")
```

:::

3. What's the default position adjustment for `geom_boxplot()`? Create a visualization of the `mpg` dataset that demonstrates it. +2pts

::: {.fold .s .o}

**Ans-1.6.1.3:**

The default position for `geom_boxplot()` is `"dodge2"`.
```{r}
# Default position (dodge2) - boxplots are side-by-side
ggplot(mpg, aes(x = class, y = hwy, color = drv)) +
  geom_boxplot()
```


The `dodge2` position places boxplots side-by-side within each class, making it easy to compare different drive types.
```{r}
# Explicitly setting position to dodge2 (same result)
ggplot(mpg, aes(x = class, y = hwy, color = drv)) +
  geom_boxplot(position = "dodge2") + geom_point()
```


**The key insight:** When you have multiple geoms with grouping (like color = drv), you need to apply the same position adjustment to all layers, or they won't align!

```{r}
ggplot(mpg, aes(x = class, y = hwy, color = drv)) +
  geom_boxplot() +  # Uses dodge2 by default
  geom_point(position = position_dodge(width = 0.75))  # Use dodge, not dodge2
```


:::

## Summary

In this chapter you learned about the layered grammar of graphics starting with aesthetics and geometries to build a simple plot, facets for splitting the plot into subsets, and position adjustments for controlling how geoms are displayed when they might overlap.

### Key concepts learned:

1. **Aesthetics**: Map variables from your data to visual properties

   - Inside `aes()` for variables: `aes(color = class)`
   
   - Outside `aes()` for constants: `color = "blue"`
   
   - **Memory trick:** Can you do `mpg$class`? YES → inside `aes()`

2. **Geoms**: The geometric objects that represent your data

   - `geom_point()`, `geom_line()`, `geom_bar()`, `geom_boxplot()`, `geom_jitter()`, etc.
   
   - Can layer multiple geoms in one plot
   
   - **Layers stack, they don't replace!**

3. **Facets**: Split your plot into subplots

   - `facet_wrap(~ variable)` for one variable
   
   - `facet_grid(rows ~ cols)` for two variables
   
   - Use `.` as placeholder: `facet_grid(. ~ cols)` or `facet_grid(rows ~ .)`

4. **Position adjustments**: Fine-tune positioning for overlapping geoms

   - `position = "stack"` - default for bars (shows totals with composition)
   
   - `position = "fill"` - proportional stacking (all bars reach 100%)
   
   - `position = "dodge"` - side-by-side bars
   
   - `position = "jitter"` - adds random noise to avoid overplotting

### The grammar of graphics works like painting:
Each `+` adds a new layer of paint on top. You can paint over something, but the old layer is still there underneath. 

**Data layers stack, but display elements (themes, labels, scales) can be modified.**

Two very useful resources for getting an overview of the complete `ggplot2` functionality are the `ggplot2` [cheatsheet](https://posit.co/resources) or in your lessons folder and the [`ggplot2` package documentation](https://ggplot2.tidyverse.org/reference).